---
title       : letterFreq!
subtitle    : A presentation for Developing Data Products
author      : D Bolotov
job         : 
framework   : io2012   # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js # {highlight.js, prettify, highlight}
hitheme     : tomorrow_edited         # zenburn tomorrow
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## letterFreq!

Ever read something and wondered what the distribution of (English) characters looks like?

Whether that wall of text contains 153 a's, or 154?

Whether your friend's phone texts are dangerously low on vowels?

What the most common English letters are?


---

## letterFreq!

Wonder no more.

LetterFreq!, a handy little app built in [Rstudio](http://www.rstudio.com/) and [Shiny](http://shiny.rstudio.com/), will take care of your letter-counting needs.

The app eats any text, dirty or clean, and shows you relative and absolute frequencies of the English letters contained therein.

Optionally, it will also plot the letter distribution from some English literature classics, or the main entries in the Oxford English dictionary, for comparison.


Let's see a demo of the histogram capabilities.

---

## letterFreq!

```{r echo=F}
require(ggplot2)

sanitize <- function(str){
    #input: raw text string
    #output: relative frequencies
    
    str <- substr(str,1,30000)
    
    newstr <- gsub("[^[:alpha:]]", "", str)
    newstr <- tolower(newstr)
    chars <- strsplit(newstr, NULL)[[1]]
    freqs <- summary(factor(chars))
    rfreqs <- freqs / sum(freqs)
    
    return(freqs)
}

fill_freqs <- function(freqs) {
    #input: frequencies (not necessarily for all 26 letters)
    #output: frequencies for all 26 letters
    
    #make 0-count vector
    alphabet <- rep(1,26)
    names(alphabet) <- letters
    
    alphabet_filled <- sapply(1:26, function(x) {freqs[names(alphabet)[x]]})
    
    alphabet_filled[is.na(alphabet_filled)] <- 0
    
    names(alphabet_filled) <- letters
    alphabet_filled <- as.data.frame(alphabet_filled)
    names(alphabet_filled) <- "rfreqs"
    
    return(alphabet_filled)
}


fplot <- function(str){
    #input: raw text, array with reference frequencies
    #output: plot
    freqs <- sanitize(str)
    rfreqs <- freqs/sum(freqs)
    
    alphabet_filled_freqs <- fill_freqs(freqs)
    alphabet_filled_rfreqs <- fill_freqs(rfreqs)
    
    alphabet_filled_new <- merge(alphabet_filled_freqs,alphabet_filled_rfreqs,by=0,all=TRUE)
    rownames(alphabet_filled_new) <- rownames(alphabet_filled_freqs)
    colnames(alphabet_filled_new) <- c("Row.names", "freqs", "rfreqs")

    p <- ggplot(data=alphabet_filled_new, aes(x=seq(1:26),y=rfreqs)) + geom_bar(stat="identity",aes(fill=rfreqs)) + 
        scale_x_discrete(breaks = c(1:26), labels=letters) +
        theme(legend.position="none") +
        geom_text(aes(label=freqs,color=freqs),size=3,hjust=0,vjust=-0.5) + 
        labs(title="Relative Letter Frequency", x = "letter", y="relative frequency")
    
    print(p)
}


```

Define a test string and get frequency counts for it:
```{r}
str <- "I am a string with numeric (123) and other non-alpha (%$^#&!) characters. 
        I live in this presentation and wonder what the rest of the world is like."
sanitize(str) #clean the string and get absolute frequency counts
```
This could use a few z's.

```{r}
str <- "I am a string with numeric (123) and other non-alpha (%$^#&!)characters. 
        I live in this presentation and wonder what the rest of the world is like. ZzZZzzZ."
sanitize(str)
```

---

## letterFreq!

Plot the distribution:
```{r echo=TRUE, fig.height=4, fig.width=6, fig.align='center'}
str <- "I am a string with numeric (123) and other non-alpha (%$^#&!)characters. 
        I live in this presentation and wonder what the rest of the world is like. ZzZZzzZ."
fplot(str)
```

Hmm. No j, but eleven i's? And not even one b? Fascinating.

Enter your own text, check other distributions, and see it in action here: [letterFreq!](https://dontpanic.shinyapps.io/letterFreq/)





