---
title       : letterFreq!
subtitle    : A presentation for Developing Data Products course
author      : D Bolotov
job         : 
framework   : io2012   # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js # {highlight.js, prettify, highlight}
hitheme     : tomorrow_edited         # zenburn tomorrow
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## letterFreq!

Ever read some text and wondered what the distribution of characters looks like?

Whether that wall of text contains 23 a's, or 24?

What the most common English letters are?


---

## letterFreq!

hey there

---

## letterFreq!

```{r echo=F}
library(ggplot2)

sanitize <- function(str){
    #input: raw text string
    #output: relative frequencies
    
    str <- substr(str,1,30000)
    
    newstr <- gsub("[^[:alpha:]]", "", str)
    newstr <- tolower(newstr)
    chars <- strsplit(newstr, NULL)[[1]]
    freqs <- summary(factor(chars))
    rfreqs <- freqs / sum(freqs)
    
    return(freqs)
}

fill_freqs <- function(freqs) {
    #input: frequencies (not necessarily for all 26 letters)
    #output: frequencies for all 26 letters
    
    #make 0-count vector
    alphabet <- rep(1,26)
    names(alphabet) <- letters
    
    alphabet_filled <- sapply(1:26, function(x) {freqs[names(alphabet)[x]]})
    
    alphabet_filled[is.na(alphabet_filled)] <- 0
    
    names(alphabet_filled) <- letters
    alphabet_filled <- as.data.frame(alphabet_filled)
    names(alphabet_filled) <- "rfreqs"
    
    return(alphabet_filled)
}


fplot <- function(str){
    #input: raw text, array with reference frequencies
    #output: plot
    freqs <- sanitize(str)
    rfreqs <- freqs/sum(freqs)
    
    alphabet_filled_freqs <- fill_freqs(freqs)
    alphabet_filled_rfreqs <- fill_freqs(rfreqs)
    
    alphabet_filled_new <- merge(alphabet_filled_freqs,alphabet_filled_rfreqs,by=0,all=TRUE)
    rownames(alphabet_filled_new) <- rownames(alphabet_filled_freqs)
    colnames(alphabet_filled_new) <- c("Row.names", "freqs", "rfreqs")

    p <- ggplot(data=alphabet_filled_new, aes(x=seq(1:26),y=rfreqs)) + geom_bar(stat="identity",aes(fill=rfreqs)) + 
        scale_x_discrete(breaks = c(1:26), labels=letters) +
        theme(legend.position="none") +
        geom_text(aes(label=freqs,color=freqs),size=3,hjust=0,vjust=-0.5) + 
        labs(title="Relative Letter Frequency", x = "letter", y="relative frequency")
    
    print(p)
}


```

Let's define a test string, get frequency counts for it, and plot them.
```{r}
str <- "I am a string with numeric (123) and other non-alpha (%$^#&!)characters."
sanitize(str) 
```



---

## letterFreq!

Let's define a test string, get frequency counts for it, and plot them.
```{r fig.height=5, fig.width=7}
fplot(str)
```

Enter your own text, check other distributions, and see it in action here: [letterFreq!](https://dontpanic.shinyapps.io/letterFreq/)





